---
title: "Generating Random Variates"
draft: false
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: STAT 3150--Statistical Computing
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(cache = FALSE, message = FALSE,
                      linewidth = 50)
set.seed(3150)
```

## Lecture Objectives

  - Recognize when to use the inverse-transform method.
  - Be able to generate random variates through transformations.
  - Derive bounding densities for accept-reject sampling.

## Motivation

  - A staple of modern statistical research is the **simulation study**.
    + Finite sample properties can then be compared to theoretical expectations.
  - More generally, by simulating data we can study the properties of a method or a model.
  - **Bayesian statistics** strongly relies on generating data to estimate the posterior density of the parameters (cf. STAT 4150).

## Inverse-Transform Method {.allowframebreaks}

  - *Recall*: Let $X$ be a random variable with CDF $F(x)$. The **quantile** function is defined as
  $$F^{-1}(p) = \inf\{x \in \mathbb{R} \mid F(x) \geq p\}.$$
  - If $X$ is continuous, this is simply the inverse function.

### Theorem

If $U$ is uniform on $[0,1]$, then $F^{-1}(U)$ has the same distribution as $X$.

  - In `R`, we can sample random variates from $U(0,1)$ by using the function `runif`:
  
```{r}
runif(5)
```

### Algorithm 

To generate random variates from $F$:

  1. Generate random variates from $U(0,1)$.
  2. Compute the quantile function $F^{-1}$.
  3. Plug-in the uniform variates into $F^{-1}$.
  
## Example {.allowframebreaks}

  - Let $X$ follow an exponential distribution with parameter $\lambda$:
  $$ F(x) = 1 - \exp(-\lambda x).$$
  - Since $X$ is continuous, the quantile function is the inverse of $F$:
  \begin{align*}
  p = 1 - \exp(-\lambda x) &\Rightarrow \exp(-\lambda x) = 1 - p\\
    &\Rightarrow -\lambda x = \log(1 - p)\\
    &\Rightarrow x = \frac{-\log(1 - p)}{\lambda}.\\
  \end{align*}

```{r}
lambda <- 1
# We want 1000 samples
n <- 1000
unif_vars <- runif(1000)
exp_vars <- -log(1 - unif_vars)/lambda
```

```{r}
# Compute theoretical quantiles
# using qexp
exp_theo <- qexp(ppoints(n))
qqplot(exp_theo, exp_vars)
# Add diagonal line
abline(a = 0, b = 1)
```

**Note**: If $U$ is uniform on $[0,1]$, so is $1 - U$.

  - Therefore $\frac{-\log(U)}{\lambda}$ also follows an $Exp(\lambda)$ distribution.
  
## Inverse Transform---Discrete Edition

```{r echo = FALSE, fig.cap="From Wikipedia"}
library(cowplot)
fig_svg <- cowplot::ggdraw() + cowplot::draw_image("../../static/images/Generalized_inversion_method.svg")
plot(fig_svg)
```

## Example {.allowframebreaks}

  - Let $X$ follow a Bernoulli distribution with parameter $p$:
  $$ F(x) = \begin{cases} 0 &\qquad x < 0,\\1 - p &\qquad x \in [0, 1),\\ 1 &\qquad x \geq 1.\end{cases}$$
  
```{r echo = FALSE, fig.width=6, fig.height=6, fig.align="center"}
plot(x = c(0, 1), y = c(0, 1), type = "n",
     xaxt = "n", xlab = "x", ylab = expression(F(x)),
     asp = 1, main = expression(p == 0.6))
axis(side = 1, at = c(0L, 1L))
segments(-1, 0, 0)
segments(0, 0.4, 1)
segments(1, 1, 2)
points(x = c(0, 1), y = c(0, 0.4), pch = 21, bg = "white")
points(x = c(0, 1), y = c(0.4, 1), pch = 19)
```

  - As we can see, we have
  $$ F^{-1}(u) = \begin{cases} 0 &\qquad u \leq 1 - p,\\ 1 &\qquad u > 1 - p.\end{cases}$$
  - In other words, we sample $U$. If it is less than $p$, we set $X = 0$; else, we set $X = 1$.
  
```{r}
p <- 0.6
n <- 1000
unif_vars <- runif(1000)
# as.numeric turns FALSE into 0 
# and TRUE into 1
bern_vars <- as.numeric(unif_vars > 1 - p)
```

```{r}
c(mean(bern_vars), var(bern_vars))
# Compare with theory
c(p, p*(1 - p))
```

## More General Transformations

  - Inverse transform is just one type of transformation!
  - We can use relationships between distributions to generate random variates. For example:
    - If $Z\sim N(0, 1)$, then $Z^2 \sim \chi^2(1)$.
    - If $V_1, \ldots, V_p \sim \chi^2(1)$, then $\sum_{i=1}^p \sim \chi^2(p)$.
    - If $U \sim \chi^2(p)$ and $V\sim\chi^2(q)$, then
    $$\frac{U/p}{V/q} \sim F(p, q).$$
    
## Example {.allowframebreaks}

```{r}
# Choose degrees of freedom
p <- 2
q <- 4

# rnorm samples from a normal distribution
U <- sum(rnorm(p)^2)
V <- sum(rnorm(q)^2)

# Take ratio
(U/p)/(V/q)
```

```{r}
# What if we want 1000 replicates?
# Use the function replicate!
# First argument: number of replicates
# Second argument: expression to be run multiple times
f_vars <- replicate(1000, {
  U <- sum(rnorm(p)^2)
  V <- sum(rnorm(q)^2)
  (U/p)/(V/q)
})
```

```{r}
qqplot(f_vars, qf(ppoints(1000), p, q))
# Add diagonal line
abline(a = 0, b = 1)
```

## Mixture of distributions {.allowframebreaks}

  - We say $X$ with distribution function $F$ follows a (discrete) **mixture of distributions** if we can write
  $$F(x) = \sum_{k=1}^K \pi_k F_k(x),$$
  where $F_k$ is a distribution function for all $k$ and $\sum_{k=1}^K \pi_k = 1$.
    + For example, the distribution of adult heights can be well-approximated by a mixture of two normal distributions with equal weights.

### Algorithm

To sample from $F = \sum_{k=1}^K \pi_k F_k$:

  1. Select a component $k$ by sampling from a multinomial distribution with probabilities $(\pi_1, \ldots, \pi_K)$.
  2. Sample from $F_k$.
  
## Example {.allowframebreaks}

  - Assume that adult heights for **male** follows a normal distribution with mean 175cm and standard deviation 7cm.
  - Assume that adult heights for **female** follows a normal distribution with mean 160cm and standard deviation 7cm.
  - Finally assume that we have an equal proportion of male and female.
  - How can we sample from the distribution of adult heights?
  
```{r}
height_vars <- replicate(1000, {
  sex <- sample(c("M", "F"), size = 1,
                prob = c(0.5, 0.5))
  if (sex == "M") {
    height <- rnorm(1, mean = 175, sd = 7)
  }
  if (sex == "F") {
    height <- rnorm(1, mean = 160, sd = 7)
  }
  return(height)
})
```

```{r}
plot(density(height_vars), ylim = c(0, 0.06))
# Add component densities
x_seq <- seq(140, 200, length.out = 100)
lines(x_seq, dnorm(x_seq, 175, 7), col = "red")
lines(x_seq, dnorm(x_seq, 160, 7), col = "green")
```

```{r}
# Sample with less code
sex <- sample(c(1, 2), size = 1000,
              replace = TRUE, 
              prob = c(0.5, 0.5))
height_vars <- rnorm(1000, c(175, 160)[sex], 7)
c(mean(height_vars), sd(height_vars))
```

## Summary

  - When we can compute the quantile function, the inverse transform is simple to implement.
    + But it can be hard to compute!
  - We can leverage relationships between distributions to transform one random variate into another.
  - **Mixture distributions** are a flexible way to model heterogeneous data, and it's "easy" to sample from them.
    + **Note**: Do not confuse *convolutions* (i.e. adding random variables) with *mixtures* (i.e. convex combination of distribution functions).

## Acceptance-Reject Method {.allowframebreaks}

  - Suppose you want to sample from a distribution $X$ with density $f$, but you can only sample from a different distribution $Y$ with density $g$.
  - Further suppose that there exists a constant $c > 1$ such that
  $$ \frac{f(t)}{g(t)} \leq c$$
  for all $t$ such that $f(t) > 0$.
  - The *Acceptance-Reject method* is a way to transform random variates of $Y$ into random variates of $X$.
  
### Algorithm

  1. Sample $y$ from $Y$.
  2. Sample a uniform variate $u$ from $U(0,1)$.
  3. Compute the ratio $r:=\frac{f(y)}{cg(y)}$. If $u < r$, set $x = y$. Otherwise, reject $y$ and repeat from Step 1.
  
**Note**: The number of iterations before we accept a draw from $Y$ follows a geometric distribution with mean $c$. So we want the constant $c$ to be as small as possible.
  
(If you want a proof of why this works, see UM Learn.)

## Example {.allowframebreaks}

  - We want to sample from $X\sim Beta(2,2)$ whose density is $f(x) = 6x(1-x)$.
    + The *proposal* distribution will be $Y\sim Beta(1,1)$ (i.e. a uniform distribution).
  - Let $t\in (0,1)$. We have
  $$\frac{f(t)}{g(t)} = \frac{6t(1-t)}{1}\leq 6,$$
  since the maximum $t$ and $1-t$ can take is 1. So we can set $c=6$.
  
```{r, echo = FALSE}
x_vec <- ppoints(100)
C <- 6
plot(x_vec, dbeta(x_vec, 2, 2), type = "l", ylim = c(0, C),
     xlab = "x", ylab = "Density")
lines(x_vec, C * dunif(x_vec))
polygon(c(0, x_vec, 1, 0), 
        c(0, dbeta(x_vec, 2, 2), 0, 0),
        col = "darkolivegreen3")
polygon(c(0, x_vec, 1, 1, 0, 0), 
        c(0, dbeta(x_vec, 2, 2), 0, C, C, 0),
        col = "indianred1")
arrows(0.78, 0, 0.78, C, code = 3)
text(c(0.5, 0.5), c(0.25, 3), labels = c("Accept", "Reject"), adj = 0.5)
```

```{r}
# Set parameters----
C <- 6 # Constant
n <- 1000 # Number of variates
k <- 0 # counter for accepted
j <- 0 # iterations
y <- numeric(n) # Allocate memory
```

```{r}
# A while loop runs until condition no longer holds
while (k < n) {
  u <- runif(1)
  j <- j + 1
  x <- runif(1) # random variate from g
  if (u < 6*x*(1-x)/C) {
    k <- k + 1
    y[k] <- x
    }
}
```

```{r}
# How many iterations did we need?
j
```

```{r}
# Compare theoretical and empirical quantiles
p <- seq(0.1, 0.9, by = 0.1)
Qhat <- quantile(y, p) # empirical
Q <- qbeta(p, 2, 2) # theoretical
```

\vspace{1cm}

```{r}
round(cbind(Qhat, Q, diff = abs(Qhat - Q)), 3)
```

 - As the graph showed, the "Rejection" region is very large.
   + In fact, it is unnecessarily large.
 - With a little bit of calculus, we can show that the maximum value of $6x(1-x)$ is 1.5.
   + In other words, we can set the constant $c = 1.5$.
   + This means that we can sample from $X$ while rejecting 4 times *less* often.

```{r, echo = FALSE}
x_vec <- ppoints(100)
C <- 1.5
plot(x_vec, dbeta(x_vec, 2, 2), type = "l", ylim = c(0, C),
     xlab = "x", ylab = "Density")
lines(x_vec, C * dunif(x_vec))
polygon(c(0, x_vec, 1, 0), 
        c(0, dbeta(x_vec, 2, 2), 0, 0),
        col = "darkolivegreen3")
polygon(c(0, x_vec, 1, 1, 0, 0), 
        c(0, dbeta(x_vec, 2, 2), 0, C, C, 0),
        col = "indianred1")
arrows(0.78, 0, 0.78, C, code = 3)
text(c(0.5, 0.1), c(0.25, 1.0), labels = c("Accept", "Reject"), adj = 0.5)
```

```{r}
C <- 1.5; k <- j <- 0 # Reset counters
while (k < n) {
  u <- runif(1)
  j <- j + 1
  x <- runif(1)
  if (u < 6*x*(1-x)/C) {
    k <- k + 1
    y[k] <- x
    }
}
```

```{r}
# How many iterations did we need this time?
j
```
