---
title: "Visualization"
draft: false
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: STAT 4690--Applied Multivariate Analysis
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Tidyverse

  - For graphics, I personally prefer using `ggplot2` than base `R` functions.
    + Of course, you're free to use whatever you prefer!
  - Therefore, I often use the `tidyverse` packages to prepare data for visualization
  - Great resources:
    + The book [*R for Data Science*](https://r4ds.had.co.nz/)
    + RStudio's [cheatsheets](https://www.rstudio.com/resources/cheatsheets/)

## Pipe operator

  - One of the important features of the `tidyverse` is the pipe operator `%>%`
  - It takes the output of a function (or of an expression) and uses it as input for the next function (or expression)

---

```{r, message = FALSE, eval = FALSE}
library(tidyverse)

count(mtcars, cyl)
# Or with the pipe
mtcars %>% count(cyl)
```

## Pipe operator

  - Note that the LHS (`mtcars`) becomes the first argument of the function appearing on the RHS (`count`)
  - In more complex examples, where multiple transformations are applied one after another, the pipe operator improves readability and avoids creating too many intermediate variables.

## Main tidyverse functions

  - `mutate`: Create a new variable as a function of the other variables
```{r eval = FALSE}
mutate(mtcars, liters_per_100km = mpg/235.215)
```
  - `filter`: Keep only rows for which some condition is `TRUE`
```{r eval = FALSE}
filter(mtcars, cyl %in% c(6, 8))
```
  - `summarise`: Apply summary function to some variables. Often used with `group_by`.
```{r eval = FALSE}
mtcars %>% group_by(cyl) %>% 
  summarise(avg_mpg = mean(mpg))
```

# Data Visualization

## Main principles

Why would we want to visualize data?

  - Quality control
  - Identify outliers
  - Find patterns of interest (EDA)
  
## Visualizing multivariate data

  - To start, you can visualize multivariate data one variable at a time.
  - Therefore, you can use the same visualizing tools you're likely familiar with.
  
## Histogram {.allowframebreaks}

```{r, message=FALSE}
library(tidyverse)
library(dslabs)

dim(olive)

olive %>% 
  ggplot(aes(oleic)) + 
  geom_histogram()

olive %>% 
  ggplot(aes(oleic, fill = region)) + 
  geom_histogram() + 
  theme(legend.position = 'top')

# Or with facets
olive_bg <- olive %>% dplyr::select(-region)
olive %>% 
  ggplot(aes(oleic, fill = region)) + 
  geom_histogram(data = olive_bg, 
                 fill = 'grey') +
  geom_histogram() +
  facet_grid(. ~ region) +
  theme(legend.position = 'top')
```

## Density plot {.allowframebreaks}

  - Another way to estimate the density is with *kernel density estimators*.
  - Let $X_1, \ldots, X_n$ be our IID sample. For $K$ a non-negative function and $h > 0$ a smoothing parameter, we have
  $$ \hat{f}_n(x) = \frac{1}{nh} \sum_{i=1}^nK\left(\frac{x - X_i}{h}\right).$$
  - Many functions $K$ can be used: gaussian, rectangular, triangular, Epanechnikov, biweight, cosine or optcosine (e.g. see [Wikipedia](https://en.wikipedia.org/wiki/Kernel_(statistics)#Kernel_functions_in_common_use))

```{r, message=FALSE}
olive %>% 
  ggplot(aes(oleic)) + 
  geom_density()

olive %>% 
  ggplot(aes(oleic, fill = region)) + 
  geom_density(alpha = 0.5) + 
  theme(legend.position = 'top')
```

## ECDF plot {.allowframebreaks}

  - Density plots are "smoothed histograms"
  - The smoothing can hide important details, or even create artifacts
  - Another way of looking at the distribution: **Empirical CDFs**
    + Easily compute/compare quantiles
    + Steepness corresponds to variance

```{r, message=FALSE}
olive %>% 
  ggplot(aes(oleic)) + 
  stat_ecdf() + 
  ylab("Cumulative Probability")

# You can add a "rug"
olive %>% 
  ggplot(aes(oleic)) + 
  stat_ecdf() + 
  geom_rug(sides = "b") + 
  ylab("Cumulative Probability")

olive %>% 
  ggplot(aes(oleic, colour = region)) + 
  stat_ecdf() + 
  ylab("Cumulative Probability") + 
  theme(legend.position = 'top')
```

## Boxplot {.allowframebreaks}

  - Box plots are a simple way to display important quantiles and identify outliers
  - Components (per Tukey):
    + A box delimiting the first and third quartile;
    + A line indicating the median;
    + Whiskers corresponding to the lowest datum still within 1.5 IQR of the lower quartile, and the highest datum still within 1.5 IQR of the upper quartile;
    + Any datum that falls outside the whiskers is considered a (potential) outlier.

```{r, message=FALSE}
olive %>% 
  ggplot(aes(y = oleic)) + 
  geom_boxplot(x = 0)

olive %>% 
  ggplot(aes(x = region, y = oleic)) + 
  geom_boxplot()

# Add all points on top of boxplots
# Note: need to remove outliers or you will get 
#       duplicates
olive %>% 
  ggplot(aes(x = region, y = oleic)) + 
  geom_boxplot(outlier.colour = NA) +
  geom_jitter(width = 0.25, height = 0)
```

# Bivariate plots

## Scatter plot {.allowframebreaks}

  - The plots above displayed information on a single variable at a time.
  - The simplest way to represent the relationship between two variables is a *scatter plot*.
  - Technically still possible with three variables, but typically more difficult to read.
  
```{r}
stars %>% 
  ggplot(aes(magnitude, temp)) + 
  geom_point()

stars %>% 
  ggplot(aes(magnitude, temp)) + 
  geom_point(aes(colour = type))
```

```{r}
library(scatterplot3d)

greenhouse_gases %>% 
  spread(gas, concentration) %>% 
  with(scatterplot3d(CH4,   # x axis
                     CO2,   # y axis
                     N2O    # z axis
))
```

## Bivariate density plot {.allowframebreaks}

```{r}
stars %>% 
  ggplot(aes(magnitude, temp)) + 
  geom_point(aes(colour = type)) +
  geom_density_2d()
```

## Bagplot {.allowframebreaks}

  - Introduced in 1999 by Rousseuw et al. as a bivariate generalization of Tukey's boxplot.
  - Help visualize location, spread, skewness, and identify potential outliers.
  - Components (details omitted):
    + The *bag*, a polygon "at the center of the data cloud" that contains at most 50% of the data points.
    + The *fence*, corresponding to an inflation of the bag (typically by a factor of 3). Observations outside the fence are potential outliers.
    + The *loop*, which is the convex hull of the non-outliers.

```{r, cache = TRUE, warning = FALSE, message = FALSE}
devtools::source_gist("00772ccea2dd0b0f1745", 
                      filename = "000_geom_bag.r")
devtools::source_gist("00772ccea2dd0b0f1745", 
                      filename = "001_bag_functions.r")

stars %>% 
  ggplot(aes(magnitude, temp)) + 
  geom_bag() + 
  theme_bw()

stars %>% 
  ggplot(aes(magnitude, temp)) + 
  geom_bag() + 
  geom_point(aes(colour = type)) +
  theme_bw()

gapminder %>%
  filter(year == 2012,
         !is.na(infant_mortality)) %>% 
  ggplot(aes(infant_mortality, life_expectancy)) + 
  geom_bag(aes(fill = continent)) + 
  geom_point(aes(colour = continent)) +
  theme_bw()

gapminder %>%
  filter(year == 2012,
         !is.na(infant_mortality)) %>% 
  ggplot(aes(infant_mortality, life_expectancy)) + 
  geom_bag(aes(fill = continent)) + 
  geom_point(aes(colour = continent)) +
  facet_wrap(~continent) +
  theme_bw()
```

# Beyond two variables

## Limitations

  - As we saw, three-dimensional scatter plots can be hard to interpret.
  - And three-dimensional bagplots would be even harder!
  - Density plots can technically be constructed for any dimension
    + But as the dimension increases, its performance *decreases* rapidly
  - **Solution**: We can look at each variable marginally and at each pairwise comparison.
  
## Pairs plot {.allowframebreaks}

  - A pairs plot arranges these univariate summaries and pairwise comparisons along a matrix.
  - Each variable corresponds to both a row and a column
  - Univariate summaries appear on the diagonal, and pairwise comparisons off the diagonal.
  - Because of symmetry, we often see a different summary of the comparison above and below the diagonal
  - I will show two packages:
    1. `GGally`
    2. `ggforce`
  
```{r, message = FALSE}
library(GGally)

olive %>% 
  dplyr::select(-region, -area) %>% 
  ggpairs
```

```{r, cache = TRUE}
library(ggforce)

olive %>% 
  dplyr::select(-region, -area) %>% 
  ggplot(aes(x = .panel_x, y = .panel_y)) + 
  geom_point() + 
  facet_matrix(vars(everything()))

olive %>% 
  dplyr::select(-region, -area) %>% 
  ggplot(aes(x = .panel_x, y = .panel_y)) + 
  geom_point() + 
  geom_autodensity() +
  facet_matrix(vars(everything()),
               layer.diag = 2)

olive %>% 
  dplyr::select(-region, -area) %>% 
  ggplot(aes(x = .panel_x, y = .panel_y)) + 
  geom_point() + 
  geom_autodensity() +
  geom_density2d() +
  facet_matrix(vars(everything()), 
               layer.diag = 2, 
               layer.upper = 3)
```

