---
title: "Visualization--Base R"
draft: false
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: STAT 3150--Statistical Computing
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Base `R`

  - Whenever I talk about "base `R`", what I mean is the functions that are part of the basic installation:
    + E.g. `base`, `stats`, and `graphics`.
  - Base `R` graphics are very powerful, but they typically require more manual tinkering to create "publication-ready" graphs.

## Main functions for data transformation {.allowframebreaks}

  - `transform`: Create a new variable as a function of the other variables
```{r eval = FALSE}
# Switch to litres per 100km
transform(mtcars, litres_per_100km = mpg/235.215)
```

\vspace{2in}

  - `subset`: Select rows or columns
```{r eval = FALSE}
# Only keep rows where cyl is equal to 6 or 8
subset(mtcars, subset = cyl %in% c(6, 8))
# Only keep cyl and mpg columns
subset(mtcars, select = c(cyl, mpg))
```

\vspace{1in}

  - `tapply`: Apply function to an array, grouping rows according to another variable.
    + **Note**: We use the operator `$` to extract the columns as vectors. 
```{r eval = TRUE, message = FALSE}
# Average mpg for each value of cyl
tapply(mtcars$mpg, mtcars$cyl, mean)
```

# Data Visualization

## Histogram {.allowframebreaks}

```{r, message=FALSE}
library(dslabs)

# Create histogram for oleic acid
hist(olive$oleic)
```

```{r hist_fill, message=FALSE, fig.keep="last"}
# Look at distribution by region
# using stacked histograms
top_hist <- hist(olive$oleic)
# Gray: North; Red: Sardinia; Blue: South
hist(olive$oleic[olive$region != "Northern Italy"],
     breaks = top_hist$breaks,
     col = "red", add = TRUE)
hist(olive$oleic[olive$region == "Southern Italy"],
     breaks = top_hist$breaks,
     col = "blue", add = TRUE)
```


```{r hist_facet, message=FALSE, eval = TRUE}
# Split plotting device in three
par(mfrow = c(1,3))
hist(olive$oleic[olive$region == "Northern Italy"])
hist(olive$oleic[olive$region == "Sardinia"])
hist(olive$oleic[olive$region == "Southern Italy"])
```

## `with` function {.allowframebreaks}

  - It can quickly become cumbersome to always use the `$` operator to extract variables.
  - The function `with` can help:
    + Inside `with`, the columns of a `data.frame` are like variables.
```{r eval = FALSE}
# Instead of this:
hist(olive$oleic)
# You can write this:
with(olive, hist(oleic))
```

```{r hist_facet2, message=FALSE, eval = TRUE}
# Same as above, but using with
par(mfrow = c(1,3))
# 2nd argument of with is an expression 
# i.e. everything between {}
with(olive, {
  hist(oleic[region == "Northern Italy"])
  hist(oleic[region == "Sardinia"])
  hist(oleic[region == "Southern Italy"])
})
```

```{r, message=FALSE, eval = FALSE}
# A more complex example
par(mfrow = c(1,3))
hist_bg <- hist(olive$oleic, col = 'white', 
                main = "Northern Italy")
with(olive, {
  hist(oleic[region == "Northern Italy"],
       col = 2, add = TRUE, breaks = hist_bg$breaks)
  plot(hist_bg, main = "Sardinia")
  hist(oleic[region == "Sardinia"],
       col = 3, add = TRUE, breaks = hist_bg$breaks)
  plot(hist_bg, main = "Southern Italy")
  hist(oleic[region == "Southern Italy"],
       col = 4, add = TRUE, breaks = hist_bg$breaks)
})
```

---

```{r hist_complex, message=FALSE, eval = TRUE, echo = FALSE}
# A more complex example
par(mfrow = c(1,3))
hist_bg <- hist(olive$oleic, col = 'white', 
                main = "Northern Italy")
with(olive, {
  hist(oleic[region == "Northern Italy"],
       col = 2, add = TRUE, breaks = hist_bg$breaks)
  plot(hist_bg, main = "Sardinia")
  hist(oleic[region == "Sardinia"],
       col = 3, add = TRUE, breaks = hist_bg$breaks)
  plot(hist_bg, main = "Southern Italy")
  hist(oleic[region == "Southern Italy"],
       col = 4, add = TRUE, breaks = hist_bg$breaks)
})
```

## Density plot {.allowframebreaks}

```{r, message=FALSE}
plot(density(olive$oleic))

# By region--first try
with(olive, {
  plot(density(oleic[region == "Northern Italy"]))
  lines(density(oleic[region == "Sardinia"]))
  lines(density(oleic[region == "Southern Italy"]))
})
```

```{r, message=FALSE}
# By region--second try
dens_ni <- with(olive, 
           density(oleic[region == "Northern Italy"]))
dens_sa <- with(olive, 
           density(oleic[region == "Sardinia"]))
dens_si <- with(olive, 
           density(oleic[region == "Southern Italy"]))

str(dens_ni)
```

```{r}
xlim <- range(c(dens_ni$x, dens_sa$x, dens_si$x))
ylim <- range(c(dens_ni$y, dens_sa$y, dens_si$y))

plot(dens_ni, xlim = xlim, ylim = ylim, col = 2)
lines(dens_sa, lty = 2, col = 3)
lines(dens_si, lty = 3, col = 4)
```

## Density plot--Comments

  - We used `plot` to get the first density plot, then we use `lines` to **add** the other density plots.
    + If we used `plot` again, it would create a new graph.
    + Also, the first call to plot determines the limits of the axes.
  - By looking at the structure of `dens_ni`, we could see that it stored the `x` and `y` values of the density estimate.
    + Therefore, we were able combine all three estimates and make sure the first plot was large enough.
  - In base `R`, we control colour with `col` (using numbers or character strings). We control the line type with `lty`.
  
## ECDF plot {.allowframebreaks}

```{r, message=FALSE}
plot(ecdf(olive$oleic))

# By region--first try
with(olive, {
  plot(ecdf(oleic[region == "Northern Italy"]))
  lines(ecdf(oleic[region == "Sardinia"]))
  lines(ecdf(oleic[region == "Southern Italy"]))
})
```

```{r, message=FALSE}
# By region--second try
ecdf_ni <- with(olive, 
                ecdf(oleic[region == "Northern Italy"]))
ecdf_sa <- with(olive,
                ecdf(oleic[region == "Sardinia"]))
ecdf_si <- with(olive, 
                ecdf(oleic[region == "Southern Italy"]))

str(ecdf_ni) # This is a function!
```

```{r}
xlim <- range(olive$oleic)

plot(ecdf_ni, xlim = xlim, col = 2)
lines(ecdf_sa, col = 3)
lines(ecdf_si, col = 4)
```

## ECDF plot--Comments

  - The output of `ecdf` is not a `matrix` or a `data.frame`, but a function!
    + In particular, it doesn't explicitly contain the `x` and `y` values, like density.
  - But we can get the bounds from the original data:
    + The `y` limits should be `c(0, 1)`, because they are probabilities.
    + The `x`-axis should cover all values in the full dataset.

## Boxplot {.allowframebreaks}

```{r, message=FALSE}
boxplot(olive$oleic)

# Split by region using a formula
boxplot(oleic ~ region, data = olive)

# Flip boxplots
boxplot(oleic ~ region, data = olive,
        horizontal = TRUE)
```

# Bivariate plots

## Scatter plot {.allowframebreaks}

```{r}
plot(stars$magnitude,
     stars$temp)

# Add colour for type of stars
with(stars, plot(magnitude, temp,
                 col = factor(type)))
```

```{r, echo = TRUE, eval = TRUE}
library(scatterplot3d)
library(tidyr)
wide_data <- spread(greenhouse_gases,
                    gas, concentration)

head(wide_data, n = 3)

with(wide_data,
     scatterplot3d(CH4,   # x axis
                   CO2,   # y axis
                   N2O    # z axis
))
```

## Scatter plot--Comments

  - Remember: for colours, `col` is either a number or a character string describing the colour (e.g. `col = "white"`)
    + The variable `type` is a character string, but not describing colours...
    + **Solution**: Transform into `factor`, which is treated as a number (i.e. the order of category).
  - We can do 3D scatterplots, but the depth of a point is hard to read.

## Bivariate density plot {.allowframebreaks}

```{r}
library(MASS)

image(kde2d(stars$magnitude,
            stars$temp))
```

## Pairs plot {.allowframebreaks}

```{r, message = FALSE}
# Select three variables
olive_sub <- subset(olive, 
                    select = c(eicosenoic, arachidic, 
                               linolenic))

plot(olive_sub)
```

```{r, eval = FALSE}
# Or alternatively:
pairs(olive_sub)
```

```{r}
# How to put histograms on the diagonal?
## From the help page for graphics::pairs
panel.hist <- function(x, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5) )
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col = "cyan",...)
}
```


```{r}
pairs(olive_sub,
      diag.panel = panel.hist)
```


## Conclusion

  - Base `R` is just as powerful as `ggplot2`.
    + But there is no consistent interface.
    + Adding components (e.g. data points on top of boxplots, histogram on pairs plots) is complex.
  - For the record, I have seen high-quality, publication-ready graphs from both approaches.
    + My personal preference is `ggplot2`, because I find it easier to iterate quickly between graphs until I find the right one.this
    
