---
title: "Visualization--ggplot"
draft: true
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: STAT 3150--Statistical Computing
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Tidyverse

  - For graphics, I personally prefer using `ggplot2` than base `R` functions.
    + Of course, you're free to use whatever you prefer!
  - Therefore, I often use the `tidyverse` packages to prepare data for visualization
  - Great resources:
    + The book [*R for Data Science*](https://r4ds.had.co.nz/)
    + RStudio's [cheatsheets](https://www.rstudio.com/resources/cheatsheets/)

## Pipe operator

  - One of the important features of the `tidyverse` is the pipe operator `%>%`
  - It takes the output of a function (or of an expression) and uses it as input for the next function (or expression)

---

```{r, message = FALSE, eval = TRUE}
library(tidyverse)

count(mtcars, cyl)
```


```{r, message = FALSE, eval = FALSE}
# Or with the pipe
# mtcars becomes the first argument of count
mtcars %>% count(cyl)
```

## Pipe operator

  - In more complex examples, with multiple function calls, the pipe operator improves readability.
```{r eval = FALSE}
# Without pipe operator
fit_model(prepare_data(dataset))
# With pipe operator
dataset %>%
  prepare_data %>% 
  fit_model
```

## Main tidyverse functions {.allowframebreaks}

  - All `tidyverse` functions take a `data.frame` as the first argument.
    + For easy piping!
  - A `data.frame` is a collection of vectors, all of the same length, but could be of different types.
    + This is the main way of organizing data in `R`.

\vspace{1in}

```{r}
# str shows the structure of an object
str(ToothGrowth)
```

\vspace{1in}

  - `mutate`: Create a new variable as a function of the other variables
```{r eval = FALSE}
# Switch to litres per 100km
mutate(mtcars, litres_per_100km = mpg/235.215)
```
  - `filter`: Keep only rows for which some condition is `TRUE`
```{r eval = FALSE}
# Only keep rows where cyl is equal to 6 or 8
filter(mtcars, cyl %in% c(6, 8))
```

\vspace{0.5in}

  - `summarise`: Apply summary function to some variables. Often used with `group_by`.
```{r eval = TRUE, message = FALSE}
# Average mpg for each value of cyl
mtcars %>% 
  group_by(cyl) %>% 
  summarise(avg_mpg = mean(mpg))
```

# Data Visualization

## Main principles

Why would we want to visualize data?

  - Quality control
  - Identify outliers
  - Find patterns of interest (EDA)
  
## Visualizing multivariate data

  - To start, you can visualize multivariate data one variable at a time.
  - Therefore, you can use the same visualizing tools you're likely familiar with.
  
## Histogram {.allowframebreaks}

```{r, message=FALSE}
library(tidyverse)
library(dslabs)

dim(olive)

# Create histogram for oleic acid
ggplot(olive,
       aes(x = oleic)) + 
  geom_histogram()
```

```{r hist_fill, message=FALSE}
# Look at distribution by region
ggplot(olive,
       aes(x = oleic, fill = region)) + 
  geom_histogram()
```


```{r hist_dodge, message=FALSE}
# Dodge instead of stack
ggplot(olive,
       aes(x = oleic, fill = region)) + 
  geom_histogram(position = "dodge")
```


```{r hist_facet, message=FALSE}
# Or with facets
ggplot(olive, 
       aes(x = oleic)) +
  geom_histogram() +
  facet_grid(. ~ region)
```

## Histogram--Summary

  - **Histograms** help visualize the distribution of a single variable.
    + It bins data and displays the counts in each bin
  - `ggplot` takes a `data.frame` as input and maps variables to different features of the graph.
    + `oleic` is mapped to the `x`-axis
    + `region` is mapped to the `fill` colour.
    + **Important**: This mapping happens inside the function `aes`.
  - `ggplot` automatically takes care of choosing the colour, drawing the limits, and printing a legend.
  - `facet_grid` can be used to display multiple plots together, one per value of the variable.

## A more complex histogram {.allowframebreaks}

```{r message = FALSE}
# Create a copy of the data to serve as background
olive_bg <- select(olive, -region)
ggplot(olive, aes(x = oleic)) + 
  # Start with grey background
  geom_histogram(data = olive_bg, 
                 fill = 'grey') +
  # Add colour on top
  geom_histogram(aes(fill = region)) +
  facet_grid(. ~ region) +
  # Move legend to top
  theme(legend.position = 'top')
```

## Density plot {.allowframebreaks}

  - Another way to estimate the density is with *kernel density estimators*.
    + See STAT 3380--Nonparametric Statistics
  - Let $X_1, \ldots, X_n$ be our IID sample. For $K$ a non-negative function and $h > 0$ a smoothing parameter, we have
  $$ \hat{f}_n(x) = \frac{1}{nh} \sum_{i=1}^nK\left(\frac{x - X_i}{h}\right).$$
  - Many functions $K$ can be used: gaussian, rectangular, triangular, Epanechnikov, biweight, cosine or optcosine (e.g. see [Wikipedia](https://en.wikipedia.org/wiki/Kernel_(statistics)#Kernel_functions_in_common_use))
    + `R` uses a Gaussian kernel by default.

```{r, message=FALSE}
ggplot(olive, aes(x = oleic)) + 
  geom_density()

# Split by region
ggplot(olive, aes(x = oleic,
                  fill = region)) + 
  geom_density()

# Add transparency
ggplot(olive, aes(x = oleic,
                  fill = region)) + 
  geom_density(alpha = 0.5)

# Alternative: stacked density plots
ggplot(olive, aes(x = oleic,
                  fill = region)) + 
  geom_density(position = "stack")
```

## Density plots--Summary

  - **Density plots** can be thought of as *smoothed* histograms.
  - We used a different *geom* to create the plot.
    + `geom_smooth` as opposed to `geom_histogram`.
  - The attribute `alpha` can be used to control transparency.
    + `alpha = 0` is completely transparent
    + `alpha = 1` is completely opaque.

## ECDF plot {.allowframebreaks}

  - The smoothing in density plots (and the binning in histograms) can hide important details, or even create artifacts
  - Another way of looking at the distribution: **Empirical CDFs**
    + Easily compute/compare quantiles
    + Steepness corresponds to variance

```{r, message=FALSE}
ggplot(olive, aes(x = oleic)) + 
  stat_ecdf()

# Change y-axis label
ggplot(olive, aes(x = oleic)) + 
  stat_ecdf() + 
  ylab("Cumulative Probability")

# You can add a "rug":
# A tick mark at the bottom for each observation
ggplot(olive, aes(x = oleic)) + 
  stat_ecdf() + 
  geom_rug(sides = "b") + 
  ylab("Cumulative Probability")

# Split by region
ggplot(olive, aes(x = oleic, 
                  colour = region)) + 
  stat_ecdf() + 
  ylab("Cumulative Probability")
```

## ECDF plot--Summary

  - ECDF plots are constructed by first summarising data:
    + This is what `stat_ecdf` does.
    + By default, the output of `stat_ecdf` is plotted using `geom_line` (see below).
  - We can combine geoms to create more complex plots
    + We added `geom_rug`.
  - For lines and points, we use the aesthetic `colour`. For areas, we use `fill`.

## Boxplot {.allowframebreaks}

  - Box plots are a simple way to display important quantiles and identify outliers
  - Components (per Tukey):
    + A box delimiting the first and third quartile;
    + A line indicating the median;
    + Whiskers corresponding to the lowest datum still within 1.5 IQR of the lower quartile, and the highest datum still within 1.5 IQR of the upper quartile;
    + Any datum that falls outside the whiskers is considered a (potential) outlier.

```{r, message=FALSE}
ggplot(olive, aes(x = oleic)) + 
  geom_boxplot(y = 0) # y = 0 is a dummy value

# Map region to y-axis
ggplot(olive, aes(x = oleic, 
                  y = region)) + 
  geom_boxplot()

# Add all points on top of boxplots
ggplot(olive, aes(x = oleic, 
                  y = region)) + 
  geom_boxplot() +
  geom_point()

# Add vertical noise to the points to reduce overlap
# Note: need to remove outliers or you will get 
#       duplicates
ggplot(olive, aes(x = oleic, 
                  y = region)) + 
  geom_boxplot(outlier.colour = NA) +
  geom_jitter(height = 0.25, width = 0) 

# Flip boxplots by switching the axes
ggplot(olive, aes(x = region, 
                  y = oleic)) + 
  geom_boxplot()
```

# Bivariate plots

## Scatter plot {.allowframebreaks}

  - The plots above displayed information on a single variable at a time.
  - The simplest way to represent the relationship between two variables is a *scatter plot*.
  - Technically still possible with three variables, but typically more difficult to read.
  
```{r}
ggplot(stars, aes(x = magnitude, 
                  y = temp)) + 
  geom_point()

# Add colour for type of stars
ggplot(stars, aes(x = magnitude, 
                  y = temp,
                  colour = type)) + 
  geom_point()
```

```{r, echo = FALSE, eval = FALSE}
library(scatterplot3d)

greenhouse_gases %>% 
  spread(gas, concentration) %>% 
  with(scatterplot3d(CH4,   # x axis
                     CO2,   # y axis
                     N2O    # z axis
))
```

## Bivariate density plot {.allowframebreaks}

```{r}
ggplot(stars, aes(x = magnitude, 
                  y = temp)) + 
  geom_density_2d()

# We can add points on top of the contour lines
ggplot(stars, aes(x = magnitude, 
                  y = temp)) + 
  geom_density_2d() +
  geom_point()

# We can colour points by star type
# Note: colour is only defined for geom_point
ggplot(stars, aes(x = magnitude, 
                  y = temp)) + 
  geom_density_2d() +
  geom_point(aes(colour = type))
```

# Beyond two variables

## Limitations

  - As we saw, three-dimensional scatter plots can be hard to interpret.
  - Density plots can technically be constructed for any dimension
    + But as the dimension increases, its performance *decreases* rapidly
  - **Solution**: We can look at each variable marginally and at each pairwise comparison.
  
## Pairs plot {.allowframebreaks}

  - A pairs plot arranges these univariate summaries and pairwise comparisons along a matrix.
  - Each variable corresponds to both a row and a column
  - Univariate summaries appear on the diagonal, and pairwise comparisons off the diagonal.
  - Because of symmetry, we often see a different summary of the comparison above and below the diagonal
  - I will show two packages:
    1. `GGally`
    2. `ggforce`
  
```{r, message = FALSE}
library(GGally)

# Select three variables
olive_sub <- olive %>% 
  select(eicosenoic, arachidic, linolenic)

ggpairs(olive_sub)
```

\vspace{0.5in}

  - As we can see, `GGally` displays the following:
    + Scatter plots below the diagonal
    + Density plots on the diagonal
    + Pearson correlations above the diagonal
  - These can all be changed---see the documentation for more information.

\vspace{0.5in}

```{r, cache = FALSE, warning=FALSE, message = FALSE}
library(ggforce)

# By default: only scatterplots
ggplot(olive_sub, aes(x = .panel_x, 
                      y = .panel_y)) + 
  geom_point() + 
  facet_matrix(vars(everything()))

# geom_autodensity adds density plots
# layer.diag = 2 restricts it to the diagonal
ggplot(olive_sub, aes(x = .panel_x, 
                      y = .panel_y)) + 
  geom_point() + 
  geom_autodensity() +
  facet_matrix(vars(everything()),
               layer.diag = 2)

# geom_density2d adds 2d density plots
# layer.diag = 3 restricts it to above the diagonal
ggplot(olive_sub, aes(x = .panel_x, 
                      y = .panel_y)) + 
  geom_point() + 
  geom_autodensity() +
  geom_density2d() +
  facet_matrix(vars(everything()), 
               layer.diag = 2, 
               layer.upper = 3)
```

  - `ggforce` is quite flexible, but the syntax is a bit confusing.
    + `x = .panel_x` and `y = .panel_y`??
    + The order of the geoms is important, because it's used in `facet_matrix` to determine where everything goes.
  - Nevertheless, the package is powerful, and it does a lot more than pairs plots!
    + `facet_zoom` to zoom in on a subset of the data on a separate panel
    + Voronoi tessalation and Delauney triangulation
    + Sina plots
    + etc.
