---
title: "Jackknife"
draft: false
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: STAT 3150--Statistical Computing
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(cache = FALSE, message = FALSE,
                      linewidth = 50)
set.seed(3150)
```

## Lecture Objective

  - Use jackknife to estimate the bias and standard error of an estimator.

## Motivation {.allowframebreaks}

  - In the previous module, we saw how we can could perform estimation and hypothesis testing using simulations.
    + **Main idea**: Simulate data from a fixed distribution, compute estimate/test statistic, and repeat the simulation to approximate the sampling distribution.
  - This approach can be very powerful when studying the behaviour of estimators, or when comparing multiple testing strategies.
  - However, there is a big obstacle in applying these methods for data analysis:
    + *They all assume we know the data generating mechanism.*
  - How can we apply these same principles for data analysis?
    + **Resampling methods**
  - We will study resampling methods for the next three modules, and we will see how they can be used for data analysis.

## Jackknife {.allowframebreaks}

  - The **jackknife** is a method that was first introduced to estimate the *bias* of an estimator.
  - We start with a sample $X_1, \ldots, X_n$. From that sample, we compute an estimate $\hat{\theta}$ of a parameter $\theta$. 
    + We are interested in estimating $E(\hat{\theta}) - \theta$.
  - For each $i$, we can also create another sample by *omitting* the $i$-th observation:
  $$X_1, \ldots,X_{i-1}, X_{i+1}, \ldots, X_n.$$
  - For each of these sample, we can also compute an estimate $\hat{\theta}_{(i)}$ of $\theta$.
    + E.g compute the sample mean or variance while omitting the $i$-th observation
  - In other words, we now have $n+1$ estimates of $\theta$!
  - The jackknife estimate of the bias $E(\hat{\theta}) - \theta$ is given by
  $$\widehat{\mathrm{bias}}_{jack} = (n-1)\left(\frac{1}{n}\sum_{i=1}^n\hat{\theta}_{(i)} - \hat{\theta}\right).$$
  
## Example {.allowframebreaks}

  - Consider the following two estimate of the variance:
  $$\hat{\sigma}^2_1 = \frac{1}{n}\sum_{i=1}^n \left(X_i - \bar{X}\right)^2, \quad \hat{\sigma}^2_2 = \frac{1}{n-1}\sum_{i=1}^n \left(X_i - \bar{X}\right)^2.$$
  - The only difference is the constant in front of the sum, which implies that $\hat{\sigma}^2_2$ is the unbiased estimate.
  - Let's compute the jackknife bias estimate of $\hat{\sigma}^2_1$.
  
```{r}
# Generate a random sample
n <- 20
xvars <- rgamma(n, shape = 3, rate = 5.5)

# Compute the estimate
theta_hat <- mean((xvars - mean(xvars))^2)
c("estimate" = theta_hat,
  "theoretical" = 3/5.5^2)
```

```{r}
# Jackknife
theta_i_hat <- numeric(n)

for (i in 1:n) {
  xvars_jack <- xvars[-i]
  mean_i <- mean(xvars_jack)
  theta_i_hat[i] <- mean((xvars_jack - mean_i)^2)
}
```

\vspace{1cm}

```{r}
# Estimate of bias
(bias <- (n-1)*(mean(theta_i_hat) - theta_hat))


c("De-biased" = theta_hat - bias,
  "Unbiased" = var(xvars))
```

## Example {.allowframebreaks}

  - Consider the `patch` dataset in the `bootstrap` package. It contains measurements of a certain hormone on the bloodstream of 8 individuals, after wearing a patch.
  - For each individual, we have three measurements: `placebo`, `oldpatch`, and `newpatch`.
  - The parameter of interest is a ratio of differences:
  $$\theta = \frac{E(\mbox{newpatch}) - E(\mbox{oldpatch})}{E(\mbox{oldpatch}) - E(\mbox{placebo})}.$$

\vspace{1cm}

```{r, linewidth = 75}
library(bootstrap)
str(patch)
```

  - `y` is `newpatch - oldpatch`, and `z` is `oldpatch - placebo`.
  - Recall that $E(X/Y) \neq E(X)/E(Y)$. So even if we have an unbiased estimate of both the numerator and the denominator of $\theta$, their ratio will generally be **biased**.
  
```{r}
# Estimate of theta
theta_hat <- mean(patch$y)/mean(patch$z)
```

```{r}
# Jackknife
n <- nrow(patch)
theta_i <- numeric(n)

for (i in 1:n) {
  theta_i[i] <- mean(patch[-i,"y"])/mean(patch[-i,"z"])
}
```

\vspace{1in}

```{r}
# Estimate of bias
(bias <- (n-1)*(mean(theta_i) - theta_hat))

c("Biased" = theta_hat,
  "De-biased" = theta_hat - bias)
```

  - The bias is significant: it represents `r scales::percent(abs(bias/theta_hat))` of the estimate.

**But be careful**: 
  
```{r}
# NOT THE SAME THING
mean(patch$y/patch$z)
```

## Estimate of the standard error

  - The jackknife can also be used to estimate the standard error of an estimate:
  $$\widehat{\mathrm{se}}_{jack} = \sqrt{\frac{n-1}{n}\sum_{i=1}^n\left(\hat{\theta}_{(i)} - \frac{1}{n}\sum_{i=1}^n\hat{\theta}_{(i)}\right)^2}.$$
  
## Example (cont'd)

```{r}
# Continuing on with the patch dataset
(se <- sqrt((n-1)*mean((theta_i - mean(theta_i))^2)))
# 95% CI
c("LB" = theta_hat - bias - 1.96*se,
  "UB" = theta_hat - bias + 1.96*se)
```

## Example {.allowframebreaks}

  - We will consider the `law` dataset in the `bootstrap` package. 
  - It contains information on average `LSAT` and `GPA` scores for 15 law schools.
  - We are interested in the correlation $\rho$ between these two variables
  
```{r}
library(bootstrap)
str(law)
```

```{r}
# Estimate of rho
(rho_hat <- cor(law$LSAT, law$GPA))
```

```{r}
# Jackknife
n <- nrow(law)
rho_i <- numeric(n)

for (i in 1:n) {
  rho_i[i] <- cor(law$LSAT[-i], law$GPA[-i])
}
```

\vspace{1cm}

```{r}
# Estimate of bias
(bias <- (n-1)*(mean(rho_i) - rho_hat))

c("Biased" = rho_hat,
  "De-biased" = rho_hat - bias)
```

```{r}
(se <- sqrt((n-1)*mean((rho_i - mean(rho_i))^2)))
# 95% CI
c("LB" = rho_hat - bias - 1.96*se,
  "UB" = rho_hat - bias + 1.96*se)
```


## Final remarks {.allowframebreaks}

  - The jackknife is a simple resampling technique to estimate bias and standard error.
    + The idea is to remove one observation at a time and recompute the estimate, so that we get a sample from the sampling distribution.
  - The theoretical details behind the jackknife are beyond the scope for this course. But two important observations:
    + The "debiased" estimate is generally only asymptotically unbiased. But its bias goes to 0 "more quickly" than the bias of the original estimator.
    + The jackknife only works well for "smooth plug-in estimators". In particular, the jackknife does **not** work well with the median.
  - The jackknife was generalized in two important ways:
    + **Bootstrap**: This will be the main topic for next week.
    + **Cross-validation**: This is a method for estimating the prediction error (see STAT 4250).
